---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, we read the raw data from the `activity.csv` file and save them to the data frame 
`acitivty`. Note that, the data in the second column `date` is reset to the `as.Date`-format.

```{r read, echo = TRUE}
unzip("activity.zip")
activity <- read.csv("activity.csv", header = TRUE)
activity[, 2] <- as.Date(activity$date)
```

The data frame looks now as follows:

```{r show, echo = TRUE}
str(activity)
```

For the plots we use `ggplot2` and `gridExtra`:

```{r ggplot, echo = TRUE}
library(ggplot2)
library(gridExtra)
```

## What is mean total number of steps taken per day?

We create a histogram of the total steps taken per day:

```{r histogram1, echo = TRUE}
StepsDay <- with(activity, tapply(steps, date, sum, na.rm = TRUE))
qplot(StepsDay,
      geom = "histogram", 
      bins = "5", 
      main = "Total Number of Steps per Day", 
      xlab = "Total Steps", 
      fill = I("blue"),
      col = I("black"),
      breaks = seq(0, 25000, by = 5000)) + 
      theme_bw(base_family = "Courier", base_size = 12)
```


```{r mean, echo = TRUE}
MeanSteps <- mean(StepsDay)
```

The mean of steps taken per day is `r MeanSteps`, 

```{r median, echo = TRUE}
MedianSteps <- median(StepsDay)
```

and the median is `r MedianSteps`. 

## What is the average daily activity pattern?
We start by looking at the time series of the 5-minute interval and the average number of 
steps taken, averaged across all days.

```{r timeseries, echo = TRUE}
AverageSteps <- with(activity, tapply(steps, interval, mean, na.rm=TRUE))
Intervals <- unique(activity$interval)
qplot(x = Intervals, 
      y = AverageSteps, 
      main = "Time Series of Average Number of Steps per Interval", 
      xlab = "Time (min)", 
      ylab = "Average Number of Steps",
      colour = "red", 
      geom = "line") + 
      theme_bw(base_family = "Courier", base_size = 12) + 
      theme(legend.position="none")
```

In order to find out which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps, we put the previously defined variables `Intervals` and 
`Average steps` in one data frame. Then, the index with the maximum number of average steps 
can easily be found. 

```{r max, echo = TRUE}
temp <- data.frame(cbind(Intervals, AverageSteps))
MaxIndex <- which.max(temp$AverageSteps)
MaxInt <- temp[MaxIndex, 1]
```

Therefore, the interval `r MaxInt` contains the maximum number of steps.

## Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). 
The presence of missing days may introduce bias into some calculations or summaries of the 
data.

```{r whichna, echo = TRUE}
IsNa <- sum(is.na(activity$steps))
```

The total number of missing values in the data set are `r IsNa`.

Now, we fill the missing values of the data set by the mean for that 5-minute interval 
averaged over all days computed above. 

```{r replace, echo = TRUE}
NaIndex <- which(is.na(activity$steps))
activityMod <- activity
for(i in 1:length(NaIndex))
{
  NaInt <- activityMod$interval[NaIndex[i]]
  activityMod[NaIndex[i], 1] <- temp[temp == NaInt, 2][1]
}
```

Therefore, the new data frame `acitivityMod` contains the same information as `activity`, but 
has no missing values.

```{r histogram2, echo = TRUE}
StepsDayMod <- with(activityMod, tapply(steps, date, sum, na.rm = TRUE))
qplot(StepsDayMod,
      geom = "histogram", 
      bins = "5", 
      main = "Total Number of Steps per Day", 
      xlab = "Total Steps", 
      fill = I("blue"),
      col = I("black"),
      breaks = seq(0, 25000, by = 5000)) + 
      theme_bw(base_family = "Courier", base_size = 12)
```

The effect of imputing the missing values is that the distribution is not anymore 
skewed towards a small number of steps per day as above. 

```{r mean2, echo = TRUE}
MeanStepsMod <- mean(StepsDayMod)
```

The mean of steps taken per day is now `r format(MeanStepsMod, digits = 7)`, 

```{r median2, echo = TRUE}
MedianStepsMod <- median(StepsDayMod)
```

and the median is `r format(MedianStepsMod, digits = 7)`. 

Both values differ from the ones computed where we simply ignored the missing values. In fact 
both values are larger than previously. Note, that mean and median are equal now.

## Are there differences in activity patterns between weekdays and weekends?

First, we create a new factor variable in the dataset with two levels "weekday" and "weekend" 
indicating whether a given date is a weekday or weekend day. We need the `dplyr` package for 
this issue.

```{r factor, echo = TRUE}
library(dplyr)
activityMod <- mutate(activityMod, 
                      day = ifelse(weekdays(activity$date) == "Saturday" |     
                                   weekdays(activity$date) == "Sunday", "weekend", 
                                   "weekday"))
activityMod$day <- as.factor(activityMod$day)
```

Now we are able to plot the average number of steps taken over the 5-minute intervals, 
averaged across all weekday days or weekend days.

```{r weekday_vs_weekend, echo = TRUE}
ActivityWeekend <- subset(activityMod, as.character(activityMod$day) == "weekend")
ActivityWeekday <- subset(activityMod, as.character(activityMod$day) == "weekday")

AverageStepsWeekend<-with(ActivityWeekend, tapply(steps, interval, mean, na.rm = TRUE))
AverageStepsWeekday<-with(ActivityWeekday, tapply(steps, interval, mean, na.rm = TRUE))

par(mfrow = c(2,1), mar = c(4,4,2,1))
p1 <- qplot(x = Intervals, 
            y = AverageStepsWeekday, 
            main = "Weekdays", 
            xlab = "Time (min)", 
            ylab = "Average Number of Steps",
            colour = "red", 
            geom = "line") + 
            theme_bw(base_family = "Courier", base_size = 10) + 
            theme(legend.position="none")
p2 <- qplot(x = Intervals, 
            y = AverageStepsWeekend, 
            main = "Weekend", 
            xlab = "Time (min)", 
            ylab = "Average Number of Steps",
            colour = "red", 
            geom = "line") + 
            theme_bw(base_family = "Courier", base_size = 10) + 
            theme(legend.position="none")
grid.arrange(p1, p2, ncol = 1, nrow = 2)
```

The average number of steps taken over weekends is larger than over weekdays on average.